Bayesian Decision Theory in Structural Geological Modeling – How Reducing Uncertainties Affects Reservoir Value Estimations

### 2018-03-09 Risk analysis/Poster/Poster STC 2017/poster.tex
Poster draft

\title{\ Summer School \enquote{CCS~in~the~High~North}}
%\author{Marco van Veen$^{1}$}
%\institute[RWTH Aachen University]{$^{1}$Institute of Computational Geoscience and Reservoir Engineering, RWTH Aachen University, Germany}
\author{Marco van Veen$^{1}$, Andrew Steadman$^{2}$, Valentin Zuchuat$^{3}$}
\institute[RWTH Aachen University]{$^{1}$Institute of Computational Geoscience and Reservoir Engineering, RWTH Aachen University, Germany\\
$^{2}$School of Earth Sciences, University of Bristol, United Kingdom\\
$^{3}$Department of Geosciences, University of Oslo, Norway
}
\date{\today}

@@

\textbf{Carbon Capture and Storage (CCS) is one of the major mitigation measures to handle climate change. Even though international organisations agreed on the importance of CCS and the effective Paris Agreement refers to the widely accepted climate scenarios, the technology is not implemented on a large scale beyond pilot projects yet, nor are economic incentives given to foster its deployment. CCS is not a cutting-edge technology, the applied methods are well-known – not only in the oil and gas industry. Buzz words like enhanced oil recovery (EOR) and carbon capture, utilisation and storage (CCUS) stir up hope to realise large-scale projects soon and in an economically viable way.}
						\vspace{0.5em}
						\begin{figure}
							\begin{minipage}{0.95\textwidth}
								\centering\includegraphics[width=1\textwidth]{figures/photos/group.jpg}
								\caption{Students of 2016 at the edge of the world's northernmost swimming pool in the abandoned Russian mining town of Pyramiden (Photo: V. Zuchuat).}
								\label{fig:group}
							\end{minipage}
						\end{figure}

The summer school \enquote{CCS in the High North} is a collaboration between the \textbf{University of Oslo} (UiO), the \textbf{Colorado School of Mines} (CSM) in Golden and the \textbf{University Centre in Svalbard} (UNIS). It offers a suite of lectures, exercises and geological field work on MSc and PhD level, which are credited with \textbf{13 ECTS}. A highly motivated group of international students with diverse backgrounds and specialisations were not only taught, but also engaged themselves in inspiring discussions.

%Unfortunately, in 2017 the course in Golden, Colorado had to be cancelled due to high political uncertainties and visa restrictions for international students.
\end{myblock}\vfill

\begin{myblock}{Course Content}

	\includegraphics[scale=0.1]{figures/icons/eco_law}~\textbf{Economy, Policies \& Risk}
	\begin{itemize}
		\item The need for CCS: Physics, history and challenges of global climate change
		\item Economics and regulations: Incentives, carbon trading, \ce{CO2} utilisation
		\item Risk Assessment: Monitoring, simulation and liability
	\end{itemize}
	\vspace{0.5em}

	\includegraphics[scale=0.1]{figures/icons/geo}~\textbf{Geology \& Reservoir}
	\begin{itemize}
		\item Trapping mechanisms \& storage options: Static vs. dynamic storage
		\item Geological reservoir characterisation: Field observations and core logging
		\item Seal properties: Mechanical and chemical compaction, global data collections and experimental studies
		\item Rock mechanics: Fractures, secondary permeability
		\item Risk of fault and fracture leakage: Sub-seismic resolution, juxtaposition, smearing
	\end{itemize}
	\vspace{0.5em}

	\includegraphics[scale=0.1]{figures/icons/tech}~\textbf{Technology}
	\begin{itemize}
		\item Physical and chemical properties of \ce{CO2}
		\item Separation technologies: Chemical absorption, phase change, polymeric membranes
		\item Enhanced Oil Recovery: Relative permeability \& mobility ratio
		\item Sequestration modelling: Rock \& fluid properties, multi-phase flow, initialisation, events/wells
	\end{itemize}

@@

\begin{myblock}{Fieldwork \& The Longyearbyen \ce{CO2} Lab}
Highlight of the Summer School were three days of fieldwork on the Arctic archipelago of Svalbard. The aim of the field investigations was to identify potential reservoir formations, evaluate and compare them on their ability to safely store \ce{CO2}. Beside two promising \ce{CO2} storage formations (see figures \ref{fig:fieldwork_adolfbukta} and \ref{fig:fieldwork_skansbukta}), also the targeted units of the Longyearbyen \ce{CO2} Lab were logged in the field close to Deltanesset. Tight sandstones with quartz-overgrowth and pore-filling clay minerals can be observed. Mudstone layers, a thrust décollement and lateral permafrost provide sufficient sealing (compare figure \ref{fig:cross_section}). Injection tests show secondary permeability caused by fractures and the sealing ability is proven by high underpressure in the reservoir \citep{braathen_longyearbyen_2012}.
	\vspace{0.5em}
	\begin{figure}
		\begin{minipage}{0.95\textwidth}
			\begin{minipage}[t]{0.32\textwidth}
				\centering\includegraphics[width=1\textwidth]{figures/photos/arrival_adolfbukta}
				\caption{Going ashore in Adolfbukta: Survival suits, rifles, redundant communications and an emergency backpack are essential HSE equipment during fieldwork on the Arctic archipelago of Svalbard (Photo: M.\,Gutierrez).}
				\label{fig:arrival_adolfbukta}
			\end{minipage}
			\hspace{\fill}
			\begin{minipage}[t]{0.32\textwidth}
				\centering\includegraphics[width=1\textwidth]{figures/photos/fieldwork_adolfbukta}
				\caption{Exposed sandstones of the Carboniferous Ebbadalen Formation north of Adolfbukta. Seven distinctive coarsening-up sequences of thick beds alternating with silty layers are expected to provide high horizontal permeability (Photo: M.\,Gutierrez).}
				\label{fig:fieldwork_adolfbukta}
			\end{minipage}
			\hspace{\fill}
			\begin{minipage}[t]{0.32\textwidth}
				\centering\includegraphics[width=1\textwidth]{figures/photos/fieldwork_skansbukta}
				\caption{Permian limestone beds of the Gipshuken and Kapp Starostin Formations are found at Skansbukta. Bioturbation structures filled with evaporitic material could provide secondary porosity. The picture shows cross-bedding in upper layers (Photo: M.\,Gutierrez).}
				\label{fig:fieldwork_skansbukta}
			\end{minipage}	
		\end{minipage}
	\end{figure}
	\vspace{0.2em}

	\begin{figure}
		\begin{minipage}{.94\textwidth}
			\centering\includegraphics[width=\textwidth]{figures/cross_section}
			\caption{Cross section of the targeted storage formation (Upper Triassic–Middle Jurassic sandstones of the Kapp Toscana Group), the overlying seal (Jurassic mudstones of the Agardfjellet Formation) and the prominent thrust décollement which is providing an additional seal. The indicated well Dh4 in the west is part of the Longyearbyen \ce{CO2} Lab, the logging took place in the east where the layers are outcropping. Modified from \cite{braathen_longyearbyen_2012}.}
			\label{fig:cross_section}
		\end{minipage}
	\end{figure}
%	\vspace{0.4em}

@@

\begin{myblock}{Limited Information \& Seal Uncertainty}

\vspace{0.3em}
\begin{figure}
	\begin{minipage}{0.95\textwidth}
		\begin{minipage}[t]{0.36\textwidth}
			\vspace{0pt}
			\includegraphics[width=\textwidth]{figures/compaction.pdf}
		\end{minipage}\hfill
		\begin{minipage}[t]{0.22\textwidth}
			\vspace{0pt}
			\caption{
			Comparison of porosity–depth trend studies and experimental results for shales and
argillaceous sediments. While for E\&P projects the seal integrity is proven by the abundance of hydrocarbons, for \ce{CO2} storage it remains uncertain. Controlling factors are clay mineral and pore fluid composition, grain sizes, pore pressure and the influence of mechanical \textbf{(MC)} and chemical compaction \textbf{(CC)}. Because core sampling of seal layers is expensive, experimental studies and global databases might be auxiliary. Still, extrapolation of experimental results has limited validity for natural anisotropy and the influence of chemical compaction \citep{mondol_experimental_2007}.
			} \label{fig:compaction}
		\end{minipage}\hfill
		\begin{minipage}[t]{0.38\textwidth}
			\vspace{0pt}
			\includegraphics[width=\textwidth]{figures/leakage.pdf}
			\caption{
			Different leakage scenarios for a \ce{CO2} storage project which are subject to uncertainty. To prevent a spill out of the storage formation \textbf{(1)}, the storage capacity must be estimated. A break of \ce{CO2} through the formation's roof \textbf{(2)} depends on seal properties and knowledge of fracture pressures (see figure \ref{fig:compaction}). Evaluation of leakage along \textbf{(3)} or through \textbf{(4)} a fault requires the analysis of juxtaposition and smearing. Sub-seismic faults represent an additional risk. Still, the highest risk of leakage exists in close proximity to wells (Figure based on lecture notes, lecture by K. Indrevær \& A. Braathen, 15.6.2017).
			} \label{fig:leakage}
		\end{minipage}
	\end{minipage}
\end{figure}

@@

\begin{myblock}{Outlook \& Motivation}
		\begin{minipage}[t]{0.72\textwidth}
			\vspace{0pt}
			Marco studies \textbf{Georesources Management} in Aachen. During his various stays in Norway -- two semesters in Stavanger and summer courses in Oslo and Longyearbyen -- he became passionate about \textbf{risk} and \textbf{decision analysis}. Currently he is writing his master thesis in the field of\textbf{ geologic modelling and uncertainty quantification} in exploration. His focus will be on the application of a \textbf{value of information (VOI)} approach. A specific project to test assumptions and models would be beneficial. The high potential of future \textbf{\ce{CO2} storage projects} provides a suitable case for application. The mentioned methods are believed to contribute to urgently needed \textbf{cost reduction} and \textbf{risk management} of CCS projects.
		\end{minipage}\hfill
		\begin{minipage}[t]{0.23\textwidth}
			\vspace{0em}
			\includegraphics[width=1\textwidth]{figures/author}
			\vspace{0.5em}
		\end{minipage}

@@

					As georesources manager who majored in geoscience I am curious about the structure and dynamics of our earth, but I like to put “these rocks” into a bigger picture. I am not only eager to understand what is going on below our feet, but also to estimate its value to society or economy and to communicate gained knowledge and accompanying risks to decision makers – in order to enable them to make good decisions.

Currently I am writing my master thesis in the field of geologic modelling and uncertainty quantification in exploration. My focus will be on the application of a value of information (VOI) approach and using open source software to demonstrate light weight solutions for high-cost projects.

I am looking for a specific project or data to test the mentioned methods on, preferably in Carbon Capture and Storage (CCS), which I believe has a great potential mitigating the dramatic consequences of climate change.

>>>>

\title{Bayesian Decision Theory in Structural Geological Modeling – How
Reducing Uncertainties Affects Reservoir Value Estimations}
%\author{Marco van Veen$^{1}$}
%\institute[RWTH Aachen University]{$^{1}$Institute of Computational Geoscience and Reservoir Engineering, RWTH Aachen University, Germany}
\author{Fabian Antonio Stamm$^{1}$, Miguel de la Varga$^{1}$, Alexander Schaaf$^{1}$, Florian Wellmann$^{1}$}
\institute[RWTH Aachen University]{$^{1}$Institute of Computational Geoscience and Reservoir Engineering, RWTH Aachen University, Germany}
\date{\today}

@@

Structural geological modeling is of central importance for the assessment of uncertain hydrocarbon accumulations in potential reservoirs. Hydrocarbon exploration and production is a high-risk, high-reward sector in which good decision making is indispensable. Actors in this field are faced with numerous uncertainties that have to be considered. We examine respective decision making from a Bayesian perspective.
%We examine respective decision making from a Bayesian perspective by considering two main approaches: (1) we treat geological modeling as a Bayesian inference problem, so that additional geological information can be incorporated as likelihood functions linked to prior parameters in a probabilistic framework; (2) we base the decision-making step on value estimation by optimizing a case-specific loss function.

\begin{myblock}{Methods}
\textbf{Construction of a 3D structural geological model:}
	\begin{itemize}
	\item Synthetic model of a simple anticline-fault trap in a potential hydrocarbon system (see Figure \ref{fig:unc_lik})
	\item Inclusion of uncertainties by assigning probability distributions to the positions of layer interfaces in depth
	\end{itemize}							
								\vspace{0.5em}
								\begin{figure}
									\begin{minipage}{0.95\textwidth}
										\centering\includegraphics[width=1\textwidth]{figures/Uncertainties_Likelihoods}
										\caption{The structural geological model illustrated as a 2D cross section (A) and a 3D voxel representation (B). The inclusion of $z$-positional uncertainties affecting layer depths and fault offset are depicted in (C) and (D). Thicknesses of the three middle layers are defined by the distances of boundary points (E) and are thus directly dependent on (C).}
										\label{fig:unc_lik}
									\end{minipage}
								\end{figure}
\textbf{Integration in a probabilistic modeling framework for Bayesian analysis (see Figure \ref{fig:model_network}):}
	\begin{itemize}
	\item Setting up a full probability model taking into account all parameter probability distributions
	\item Bayesian updating: Conditioning of parameters on additionally observed data via likelihood functions
	\item Approximation of posterior distributions through the use of MCMC sampling
	\end{itemize}
\textbf{Evaluation of modeling results:}
	\begin{itemize}
	\item Shannon entropy for uncertainty visualization (after \citet{wellmann2012uncertainties}.
	\item Implementation of algorithms for structural analysis, trap recognition (see Figure \ref{fig:trap_cond}) and calculation of recoverable oil volumes (ROV)
	\item Decision making based on the optimization of a case-specific custom loss function that considers differently risk-affine actors
	\end{itemize}

	%We examine respective decision making from a Bayesian perspective by considering two main
	%approaches: (1) we treat geological modeling as a Bayesian inference problem, so that additional geological
	%information can be incorporated as likelihood functions linked to prior parameters in a probabilistic framework; (2)
	%we base the decision-making step on value estimation by optimizing a case-specific loss function. This function is
	%customized to reflect the decision making of differently risk-affine actors relative to previously computed reservoir
	%value probability distributions.
	%We apply our approach to synthetic geological models which are constructed to represent potential hydro-
	%carbon systems. Markov chain Monte Carlo sampling is used to approximate posterior models of reduced
	%uncertainty.

	%For the valuation of prior and posterior models, we also develop algorithms for automatic trap
	%	recognition and volume calculation.

							\vspace{0.5em}
							\begin{figure}
								\begin{minipage}{0.8\textwidth}
									\centering\includegraphics[width=1\textwidth]{figures/Trap_Cond_H}
									\caption{To be recognized as part of a trap, a reservoir voxel (1) has
									to be positioned in the	footwall (2). Trap closure is defined by the seal shape and the normal fault (3). The maximum trap fill is defined by either the anticlinal spill point
									(3;a) or a point of leakage across the fault, depending on juxtapositions with layers
									underlying (3;b) or overlying the seal (3;c). The latter is only relevant if the
									critical Shale Smear Factor (SSF$_c$) is exceeded, as determined over D and T in (4).}
									\label{fig:trap_cond}
								\end{minipage}
							\end{figure}

@@

\begin{myblock}{Results}
		\begin{minipage}[t]{0.48\textwidth}
					\vspace{0pt}
					Posterior ROV probability distributions are realized depending on the nature of likelihoods implemented in the Bayesian inference step. Applying the custom loss function shows that the various Bayes estimators shift according to the characteristics of this underlying	value distribution. While bimodality and overall uncertainty lead to separation, risk-averse and risk-friendly decisions converge and decrease in expected loss given narrower unimodal distributions. A decisive factor in our model is the reliability of the sealing across the fault, as it defines the probability of complete trap failure. Two examples are summarized in Figures \ref{fig:ML4} and \ref{fig:ML5}.
				\end{minipage}\hfill
				\begin{minipage}[t]{0.48\textwidth}
					\vspace{0em}
					\begin{figure}
					\centering
					\includegraphics[width=1\textwidth]{figures/model_network}
					\caption{Illustration of the probabilistic model as a hierarchical Bayesian network after \citet{koller2009probabilistic}. Stochastic nodes are represented by ellipses (blue for
					priors, yellow for likelihoods). Deterministic functions are depicted as triangles in
					red. Arrows indicate direct connections from parent to child nodes.}
					\label{fig:model_network}
					\end{figure}
					\vspace{0.5em}
				\end{minipage}

	\vspace{0.5em}
	\begin{figure}
		\begin{minipage}{0.98\textwidth}
			\begin{minipage}[t]{0.5\textwidth}
				\centering\includegraphics[width=1\textwidth]{figures/ML4}
				\caption{Evaluation of a model conditioned on thickness and SSF likelihoods which assure a reliable trap closure and reinforce the probability of a high ROV. Shannon entropy was significantly reduced in the posterior model (b). Cross-fault leakage to overlying layers was eliminated as a trap control mechanism (c) and coincides with the disappearance of the probability of failure in the posterior ROV distribution (d). The respective loss functions are plotted in (e), showing high decision convergence.}
				\label{fig:ML4}
			\end{minipage}
			\hspace{\fill}
			\begin{minipage}[t]{0.5\textwidth}
				\centering\includegraphics[width=1\textwidth]{figures/ML5}
				\caption{Evaluation of a model conditioned on a SSF likelihood which narrows around the critical SSF$_c$ and thus amplifies the duality in the posterior ROV distribution (d). Shannon entropies are only slightly diminished (a,b). general cross-fault leakage remains a significant trap control mechanism (c). The resulting loss function realizations show a high divergence of differently risk-affine decisions.}
				\label{fig:ML5}
			\end{minipage}	
		\end{minipage}
	\end{figure}
	\vspace{0.2em}

%	\vspace{0.5em}
%	\begin{figure}
%		\begin{minipage}{0.95\textwidth}
%			\begin{minipage}[t]{0.32\textwidth}
%				\centering\includegraphics[width=1\textwidth]{figures/ML2}
%				\caption{...}
%				\label{fig:ML2}
%			\end{minipage}
%			\hspace{\fill}
%			\begin{minipage}[t]{0.32\textwidth}
%				\centering\includegraphics[width=1\textwidth]{figures/ML4}
%				\caption{....}
%				\label{fig:ML4}
%			\end{minipage}
%			\hspace{\fill}
%			\begin{minipage}[t]{0.32\textwidth}
%				\centering\includegraphics[width=1\textwidth]{figures/ML5}
%				\caption{...}
%				\label{fig:ML5}
%			\end{minipage}	
%		\end{minipage}
%	\end{figure}
%	\vspace{0.2em}

	%\begin{figure}
	%	\begin{minipage}{.94\textwidth}
	%		\centering\includegraphics[width=\textwidth]{figures/cross_section}
	%		\caption{...}
	%		\label{fig:cross_section}
	%	\end{minipage}
	%\end{figure}
%	\vspace{0.4em}

@@

\begin{myblock}{Conclusions}
\begin{itemize}
\item The degree of decision convergence can be considered a measure for the state of knowledge and its inherent uncertainty at the moment of decision making.
\item This decisive uncertainty does not change in alignment with model uncertainty but depends on alterations of critical parameters and respective interdependencies.
\item Actors are affected differently by one set of information, depending on their risk affinity.
\item It is important to identify the model parameters which are most influential for the final decision in order to
optimize the decision-making process.
\end{itemize}
These results and conclusions refer to a generic hydrocarbon system case study but are transferable to other fields where decisions
are based on uncertain geological models, for example in hydrogeological or geothermal exploration.
%\vspace{0.3em}
%\begin{figure}
%	\begin{minipage}{0.95\textwidth}
%		\begin{minipage}[t]{0.36\textwidth}
%			\vspace{0pt}
%			\includegraphics[width=\textwidth]{figures/compaction.pdf}
%		\end{minipage}\hfill
%		\begin{minipage}[t]{0.22\textwidth}
%			\vspace{0pt}
%			\caption{
%			Comparison of porosity–depth trend studies and experimental results for shales and
%argillaceous sediments. While for E\&P projects the seal integrity is proven by the abundance of hydrocarbons, for \ce{CO2} storage it remains uncertain. Controlling factors are clay mineral and pore fluid composition, grain sizes, pore pressure and the influence of mechanical \textbf{(MC)} and chemical compaction \textbf{(CC)}. Because core sampling of seal layers is expensive, experimental studies and global databases might be auxiliary. Still, extrapolation of experimental results has limited validity for natural anisotropy and the influence of chemical compaction \citep{mondol_experimental_2007}.
%			} \label{fig:compaction}
%		\end{minipage}\hfill
%		\begin{minipage}[t]{0.38\textwidth}
%			\vspace{0pt}
%			\includegraphics[width=\textwidth]{figures/leakage.pdf}
%			\caption{
%			Different leakage scenarios for a \ce{CO2} storage project which are subject to uncertainty. To prevent a spill out of the storage formation \textbf{(1)}, the storage capacity must be estimated. A break of \ce{CO2} through the formation's roof \textbf{(2)} depends on seal properties and knowledge of fracture pressures (see figure \ref{fig:compaction}). Evaluation of leakage along \textbf{(3)} or through \textbf{(4)} a fault requires the analysis of juxtaposition and smearing. Sub-seismic faults represent an additional risk. Still, the highest risk of leakage exists in close proximity to wells (Figure based on lecture notes, lecture by K. Indrevær \& A. Braathen, 15.6.2017).
%			} \label{fig:leakage}
%		\end{minipage}
%	\end{minipage}
%\end{figure}

@@

\begin{myblock}{Acknowledgements}
This research poster is based on a master thesis submitted by Fabian Antonio Stamm to the Department of Computational Geoscience and Reservoir Engineering (CGRE) at the RWTH Aachen University in 2017. 3D geological models were created using GemPy \citep{gmd-2018-61} and integrated into a probabilistic framework using PyMC \citep{salvatier2016pymc3}.

		%\begin{minipage}[t]{0.72\textwidth}
		%	\vspace{0pt}
		%	This research poster is based on work conducted as part of a master thesis submitted to the Institute of Computational Geoscience and Reservoir Engineering (CGRE) at the RWTH Aachen University in 2017.
		%	
		%	\begin{itemize}
		%	\item E-mail: fabian.stamm@rwth-aachen.de
		%	\item Phone: + 49 174 5231499
		%	\end{itemize}

		%\end{minipage}\hfill
		%\begin{minipage}[t]{0.23\textwidth}
		%	\vspace{0em}
		%	\includegraphics[width=1\textwidth]{figures/author}
		%	\vspace{0.5em}
		%\end{minipage}

@@
